The Economics of Wildfire Aviation

### 2023-10-28 manuscript/manuscript.qmd
updated ms with some info about external costs of retardant

The procurement of aviation resources influences federal expenditure on firefighting aviation. For nearly the past two decades, federal firefighting agencies have contracted with private companies that own and maintain firefighting aircraft. Currently, federal agencies either lease aircraft for much of the season or sign short-term contracts for a higher daily cost. The structure and prices of these contracts and the procurement mechanism influence how aviation resources are used and the costs borne by firefighting agencies.

The objective of this paper is to describe and analyze two economic aspects of aviation used for wildfire response. First, we apply economic principles to the procurement of aviation resources and discuss how and where economic efficiency can be gained. Second, we discuss how the benefits of aviation are not well measured or tracked, making it difficult to analyze the benefits relative to the costs. While this paper will not come to clear conclusions about the use of aviation, it will frame the economic problems and attempt to provide a path forward for how to solve them.

@@

Cost benefit analysis requires the accounting of the costs and benefits of an intervention over a relevant time period [@boardman2018]. The direct cost of using aviation to fight fire is recorded because the government is responsible for payment.[^4] The benefits of aviation are harder to quantify because they involve the avoided costs and losses of fires that were prevented from growing and damaging valued assets. If aviation resources reduce the probability that a fire spreads and damages property, we do not observe the counterfactual - what would have happened in the absence of aviation use. This section describes how aviation resources are currently used to fight fire, discusses the conceptual challenges to measuring benefits, and proposes data driven approaches to measuring the benefits of aviation given currently available data.

[^4]: The indirect costs of aviation include the risk to pilots that fly in dangerous conditions around challenging terrain. @butler2015 document that 78 of the 298 (26%) recorded wildland firefighting fatalities were attributable to aviation incidents between 2000 and 2013.

>>>>

The procurement of aviation resources influences federal expenditure on firefighting aviation. For nearly the past two decades, federal firefighting agencies have contracted with private companies that own and maintain firefighting aircraft. Currently, federal agencies lease and own aircraft. In the case of leased aircraft, contracts are structured in one of two ways: a near season long exclusive use agreement or short-term contracts that generally caries a price premium. The structure and prices of these contracts and the procurement mechanism influence how aviation resources are used and the costs borne by firefighting agencies.

The objective of this paper is to describe and analyze two economic aspects of aviation used for wildfire response. First, we apply economic principles to the procurement of aviation resources and discuss how federal agencies may improve economic efficiency. Second, we discuss how the benefits of aviation are not well measured or tracked, making it difficult to conduct benefit cost analysis. While this paper will not come to clear conclusions about the use of aviation, it will frame the economic problems and attempt to provide a path forward for how to solve them.

@@

Cost benefit analysis requires the accounting of the costs and benefits of an intervention over a relevant time period [@boardman2018]. The financial cost of using aviation to fight fire is recorded because the government is responsible for payment.[^4] The benefits of aviation are harder to quantify because they involve the avoided costs and losses of fires that were prevented from growing and damaging valued assets. If aviation resources reduce the probability that a fire spreads and damages property, we do not observe the counterfactual - what would have happened in the absence of aviation use. This section describes how aviation resources are currently used to fight fire, discusses the conceptual challenges to measuring benefits, and proposes data driven approaches to measuring the benefits of aviation given currently available data.

[^4]: There are other costs of using aviation, including the risk to pilots that fly in dangerous conditions around challenging terrain and the external costs of the retardant. @butler2015 document that 78 of the 298 (26%) recorded wildland firefighting fatalities were attributable to aviation incidents between 2000 and 2013. The external costs of applying retardant are not well studied, but they are the subject of a current court case [@brown2023].
